A conditional permutation-based approach to test confounder effect and center-bias in machine learning models

### 2021-08-15 manuscript.tex
polished (inner review)

Both acquisition batch and age group were statistically significantly associated with fluid intelligence ($R^2=0.032$ and $0.011$ and $p<0.001$ and $=0.001$, respectively, see also Table \ref{tab:unconditional-pvals}). The model trained on the raw (unadjusted) connectivity features predicted fluid intelligence with a medium effect size ($R^2=0.095$, $p<0.001$).
The partial confounder test revealed that the model was significantly biased both by age group and acquisition batch (both $p<0.0001$, first column of Fig. \ref{fig:hcp}) with later phases of the acquisition and lower age being weakly-moderately ($R^2=0.084$ and $0.041$, respectively) associated to larger predicted values.

After applying confound mitigation approaches (feature regression or COMBAT) the partial confound test did not provide evidence of confounder bias anymore ($p > 0.05$ for all, second and third columns of Fig. \ref{fig:hcp}). Both feature regression and COMBAT decreased the predictive performance in the case of both confounders, with COMBAT providing better performances ($R^2=0.75$ and $0.76$ when applied to remove the effect of acquisition and age, respectively).

@@

Imaging center and in-scanner motion (normalized mean framewise displacement) were statistically significantly associated with ASD diagnosis ($R^2=0.019$ and $0.028$, respectively, $p<0.001$ for both, see also Table \ref{tab:unconditional-pvals}). The model trained on the raw (unadjusted) connectivity features predicted diagnosis with a medium effect size ($R^2=0.126$, $ROC AUC = 0.71$, $p<0.001$).
The partial confounder test revealed that the raw model was significantly biased both for age group and acquisition batch (both $p<0.0001$, see first column on Fig. \ref{fig:abide}). Several sites (e.g. Carnegine Mellon University, University of Leuven, Social Brain Lab UMC Groningen) and lower motion being weakly-moderately ($R^2=0.071$ and $0.034$, respectively) associated to a lower probability for the participant to be assigned to the ASD group by the model.

Both feature regression and COMBAT seemed to significantly attenuate the effect of center, however the partial confound test still provided evidence for a significant center-bias ($p < 0.04$ and $0.01$ for feature regression and COMBAT, respectively, see the second and third columns of Fig \ref{fig:abide}).

When trying to mitigate the effect of in-scanner motion (bottom row on Fig \ref{fig:abide}), feature regression obviously failed to remove the motion-bias from the model and, in fact, it introduced a paradoxical inverse dependence of the predictions on motion estimates. The partial confounder successfully detected this strong bias ($p<0.0001$). Applying COMBAT to remove the effect of motion (by using binned motion estimates), in turn, seemed to effectively mitigate motion-bias, as suggested by the partial confound test ($p=0.64$, bottom right panel of Fig \ref{fig:abide}).

Both feature regression and COMBAT slightly improved the predictive performance when mitigating center-effects ($R^2=0.111$ and $0.132$ and $AUC=0.76$ and $0.75$, respectively). With both feature regression and COMBAT, the effort of mitigating motion effects happened at the cost of a drop in predictive performance ($R^2=0.02$ and $0.111$ and $AUC=0.59$ and $0.70$, respectively)

@@

% tab:unconditional-pvals
\renewcommand{\arraystretch}{1.2}
\begin{table}
\centering
\begin{tabular}{lll|ll|ll|ll|ll} 
dataset & conf. & method & $R^2_{y, c}$ & $p_{y, c}$ & $R^2_{\hat{y}, c}$ & $p_{\hat{y}, c}$ & $R^2_{\hat{y}, y}$ & $p_{\hat{y}, y}$ & partial test & full test  \\
\hline
HCP & acq.  & raw      & 0.032 & <0.001 & 0.084 & <0.001  & 0.095 & <0.001 & <0.001 & <0.001 \\
    &              & f.reg.    & & &0.003 & 0.99 & 0.066  & <0.001  & 0.99 & <0.001 \\
    &              & COMBAT    & & &0.008 & 0.75 & 0.075  & <0.001  & 0.88 & <0.001\\
\hline
    & age   & raw       & 0.011 & 0.001  & 0.041 & <0.001  & 0.095  & <0.001 & <0.001 & <0.001 \\
    &       & f.reg.    & & &0.001 & 0.3 & 0.07  & <0.001  & 0.5 & <0.001 \\
    &       & COMBAT    & & &0.004 & 0.055 & 0.076 & <0.001 & 0.12 & <0.001 \\
\hline
ABIDE   & center   & raw       & 0.019  & <0.001 &  0.169 & <0.001& 0.126     & <0.001 & <0.001 & <0.001 \\
        &          & f.reg.    &  & &  0.042 & 0.007 & 0.173     & <0.001 & 0.04 & <0.001 \\
        &          & COMBAT    &  & &  0.05 & 0.001 & 0.17     & <0.001 & 0.01 & <0.001 \\
\hline
        & motion   & raw       & 0.028 & <0.001 & 0.111    &  <0.001 & 0.126    & <0.001 & <0.001& <0.001 \\
        &          & f.reg.    & & & 0.359    & <0.001  & 0.02   & <0.001 & <0.001 & 0.09 \\
        &          & COMBAT    & & &  0.002 & 0.202 & 0.111     & <0.001 & 0.64 & <0.001 \\

@@

Confounder-bias can significantly harm the generalizability and biomedical validity of predictive models. Testing confounder-bias of predictive models can be achieved by considering the "confounder-problem" as assessing \emph{conditional independence} between the predicted values and either the target or the confounder variable. These two scenarios result in two different test, testing whether the model is partially or fully driven by the given confounder.
Importantly, in predictive modelling, the model predictions often follow a non-normal distribution, thus any test for confounder bias must be, at least partly, non-parametric.
However, as shown by \citet{shah2020hardness} in their 'no free lunch' theorem, it is effectively impossible to establish a fully non-parametric conditional independence test with a valid type I error control and non-trivial power.
Here I propose the use of the conditional permutation test (CPT) of \cite{berrett2020conditional} to test confounder bias in predictive models. Adapting this approach for the confounder testing problem makes it possible to place assumptions about the joint distribution of only two out of the three variables involved; that is, it allows constructing both the partial and full confounder tests so that no assumption is made about the distribution of the model output. 
The tests are made available in the python package \emph{mlconfound} which builds upon a parallelized, computationally effective implementation of the Markov-chain Monte-Carlo based parallel pairwise sampler of \cite{berrett2020conditional}. It utilizes, coefficient of determination ($R^2$ or pseudo $R^2$ in case of categorical confounder or classification) as a test statistic and assumes that the conditional distribution between the target variable and the confounder is normal. The partial and full confounder tests available in \emph{mlconfound} can be uniformly applied for any predictive model, given that realistic (e.g. cross-validated) model predictions can be obtained. Both tests require only the target variable, the model predictions and the putative confounder as input, thus, they can be performed with a negligible extra computational cost; re-fitting the model is not required.

As expected from theoretical justification, both tests displayed a valid type I error control in the simulation with the necessary assumptions, that is, even if $\yhat$ is not normally distributed. In fact, the tests were found to become somewhat conservative (i.e. lower than expected type I error rates ) in the case of high-performance predictive models (when the $R^2$ between the predicted and observed values is greater than 0.7).

@@

Next to their statistical power, the practical applicability of the proposed tests is also highly affected by their robustness of the normality assumptions on the conditional distribution of $\y$ and $\c$. Importantly, the simulations strongly suggest that both tests are strikingly robust to the violation of this assumption. Type I error is well controlled even in case of highly skewed and heavy tailed distributions (kurtosis > 5, skewness > 2) and gets significantly inflated only with extreme violations of normality (see the purple distribution on Fig. \ref{fig:sim-non-normal}). Due to the conservatives of the tests, if the predictive model is very accurate, even this extremely skewed and heavy tailed distribution resulted in lower than the nominal type I error rates.

@@

Validity, power and robustness renders the proposed tests as unprecedented model diagnostic tools in a wide spectrum of predictive modelling scenarios.
For instance, in biomedical research, confounder-bias has often been investigated (if investigated at all) by testing the (unconditioned) association between the confounder variable and the predictive values, see e.g. \citep{spisak2020pain}. However, the significance of this association does not necessarily imply a significant confound-bias of the model, especially if the confounder is also associated to the true target variable. In this case, namely, the model still might not be directly driven by the confounder and the dependence of the predictions on the confounder might be only marginal, to be explained solely by the confounder-target association. 
Two examples for this situation can be found in the presented analysis of real data; when applying COMBAT to mitigate age-bias in the HCP data and when regressing out the center-effect from the ABIDE data (see Table \ref{tab:unconditional-pvals}).

Arguably, in certain applications, the researcher might aim to fully eliminate confounder-related variance of the predictions and in this case testing the association (unconditional independence) between the confounder and the predictions might be sufficient. However, a much more realistic aim is to construct models that allow a reasonable degree of marginal confounder-association of the predictions, but are likely not directly driven by confounder-information, i.e. the prediction-confounder association is proportional to, and can likely be fully explained by, the target-confounder association.
Put simply, this is what is tested by the proposed partial confounder test.

Obviously, a true confounder bias might still be present in the model, even if the the target-confounder association likely explains the prediction-confounder association, i.e. the model might still use the confounder's effect on the features (e.g artifacts) to construct the corresponding part of the variance in the prediction. Nevertheless, in most of the cases, it does not seem likely that the model ,at the same time, (i) misses that specific "part" of the target variable's variance in the features that is shared with the confounder (ii) captures exclusively those components of the confounder's variance in the features that is needed to mimic an unbiased model.
Usually a much more plausible explanation is that a direct confounder effect in the model output is simply negligible  model (e.g because it is not to be found in the features after confound mitigation).

@@

To demonstrate the usefulness of the proposed tests in detecting various types of confounder-bias, here they have been deployed in two research settings where specific types of confounding effects have previously been described. 
First, functional connectivity data from the Human Connectome Project (HCP) \citep{van2013wu} was used to build predictive models of fluid intelligence and to test for the previously discussed confounding effect of age \citep{lohmann2021predicting, dubois2018distributed} and,  additionally, the - somewhat underdiscussed - batch-like effect of acquisition date of the data within the course of the data acquisition process.
Second, functional connectivity data from the ABIDE \citep{di2014autism} database was used to investigate the potential motion- and center-bias (as previously reported by e.g. \cite{spisak2014voxel, spisak2019optimal, gotts2013perils}) when training models that aim to predict ASD diagnosis.

@@

In case of the HCP dataset, the weak but statistically significant association between age and IQ predictions could likely exaggerate to a serious bias when testing the model on data of participants outside of the (relatively narrow)
age range of the HCP sample. In this case, the bias would likely significantly harm the out-of-sample generalizability of this model. The bias of the same model to the acquisition batch (phase of the study the acquisition belongs to) can also be problematic and it has not yet been thoroughly discussed in case of the HCP dataset. There can be manifold reasons for the observed associations. Fluid intelligence of the participants can be affected by a changing selection bias during participant recruitment (e.g. as a consequence of the human connectome project receiving an increasing degree of public interest during its course, the sociodemographic and educational characteristics of the volunteer sample might have changed). To explain the associated change in the target variable, the model might utilise changing characteristics of the data (caused by e.g. slight changes in the calibration settings of the measuring equipment or the changing level of experience of the research staff). This is obviously undesirable for any brain-based predictive model of fluid intelligence.

In the ABIDE dataset, neither the medium-strong center-bias nor the age-bias is surprising but both are obviously problematic for a diagnostic model of autism. For instance the model trained on the raw features - depending on the calibration of the predicted class probabilities - might classify all participants form e.g. the CMU (Carnegie Mellon University) center as neurotypical control participants. It's easy to see that this bias can possibly lead to a very low predictive performance on out-of-center data. Similarly, a model biased by motion - next to having questionable neuroscientific validity - might systematically fail in populations with higher motion or altered motor characteristics (as known e.g. for ADHD \citep{eloyan2012automated}).

Confounder mitigation techniques, with one exception, tended to decrease confounder bias. The partial confounder test provided detailed insights into the effectiveness of the techniques for each confounder. It revealed that, in the case of both the acquisition-bias and the age-bias was very effectively removed by both feature regression and COMBAT (p>0.05 for all). Given the high power of the test at these sample sizes (N=999 and 866 for the HCP and the ABIDE datasets, respectively), any remaining confounder-bias is most probably very safely negligible and well out of the range of practical relevance.

On the other hand, the partial confoudner test also showed, that the performance of the investigated confound mitigation techniques is much less convincing for the ABIDE dataset. The center-bias of the classification in this massively multi-center dataset was, namely, found to be significant, even after feature regression and COMBAT. Nevertheless, the p-values returned by the partial confounder test were larger with one order of magnitude than without any confound mitigation, suggesting that a large part of the center effects was still successfully removed by both approaches. While determining the relevance of the remaining bias is to be performed based on the experimental context and the requirements towards the predictive marker, significant p-values of the partial confounder test generally suggest that these models would either require more effective data harmonization or a validation on external data, that confirms the low practical relevance of the amount of bias.

Motion-bias in the ABIDE dataset was not detectable anymore after COMBAT, but remained highly significant after feature regression.
Interestingly, in this case, the association between the confounder and the predictions become even stronger than before but with an opposite sign. Meanwhile, the prediction performance significantly dropped (but still remained statistically significant). 
A possible explanation to these phenomena is that in-scanner motion, as previously described \citep{fournier2010motor, anzulewicz2016toward}, has manifold links to ASD. Therefore, regressing out motion estimates from the connectivity features might eliminate significant amount of signal-of-interest (especially in functional connections that are associated with motor deficits in ASD).
On the other hand, motion artefacts were previously shown to display complex spatiotemporal patterns in fMRI-based functional connectivity estimates \cite{spisak2014voxel} and simply regressing them out of the features is far from being sufficient for removing their effect on the model. The so-called "residual motion artefacts" and their regional interactions, as described in the referenced work, might be responsible for the paradox behavior of this model, which raises caution for using feature regression to mitigate the effect of motion artifacts and similarly complex confounders.

@@

 In sum, the application of the partial confounder test suggests that confounder-bias should be much more carefully investigated and reported in the predictive modelling literature as (i) variables as trivial as the date of the acquisition can cause significant model-bias and (ii) in certain situations, a sufficient mitigation of confounder-bias requires more effective solutions than feature regression and COMBAT. The partial confoudner test can be considered as a useful benchmark to guide the search for a suitable confounder mitigation approach.

While the real data examples in this paper were not typical cases of the potential application of the full confounder test, it has still provided important insights into model biases. Namely, in the case of the aforementioned paradox motion-regression model in the ABIDE dataset, the full confoudner test did not provide any evidence that the model captures any extra variance over that explained by the confounder. In other words, the test indicated that this model might be, in fact, exclusively driven by motion artifacts.
For all other models, the full confound test confirmed that - even if there was a significant confounder-bias - the model was not fully driven by the confounder (in other words, the target and the prediction were not independent, given the confounders).
This examples highlights that the full confounder test might become useful in exploratory phases of model development, where models might be still severely biased and the questions is whether there is any biomedical relevant signal captured by the model, at all.

@@

The lack of rigorous statistical tests for confounder-bias significantly hampers the development of predictive models in many fields of research. Even tough several approaches for mitigating the effects of counfounder variables in machine learning have been recently proposed, these can potentially remove signal-of-interest together with the confounder signal and an informed decision about confound mitigation strategy is impossible without a proper test of confounder-bias in the resulting model.

To fill this critical gap in predictive model development, here I proposed two novel tests, the partial and the full confounder tests, which probe the null hypotheses of 'no confounder-bias' and 'full confounder-bias', respectively. Both approaches are based on a recently published framework for conditional permutation testing, with a mathematical justification of valid type I error control. The tests were constructed so that they do not place any assumption on the distribution of the model predictions and require only the target variable, the model predictions and the putative confounder as input, that is, they can be performed very efficiently, without needing to re-fit the model.

Given their simplicity, high statistical power, robustness to the violation of the assumptions and thanks to the computationally effective implementation (python package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}}), the tests are deployable in a wide variety of applications.

As demonstrated on functional brain connectivity-based predictive models of fluid intelligence and autism, the tests can guide the optimization of the confound mitigation strategy and significantly enhances the quantitative assessment of the robustness, generalizability and neurobiological validity of predictive models in biomedical research, thereby paving the way for the development of clinically useful machine learning biomarkers.

>>>>

Both acquisition batch and age group were statistically significantly associated with fluid intelligence ($R^2=0.032$ and $0.011$ and $p<0.001$ and $=0.001$, respectively, see also Table \ref{tab:unconditional-pvals}). The model trained on the raw (unadjusted) connectivity features predicted fluid intelligence with a medium effect size ($R^2=0.095$, $p<0.001$).
The partial confounder test revealed that the model was significantly biased both by age group and acquisition batch (both $p<0.0001$, first column of Fig. \ref{fig:hcp}) with later phases of the acquisition and lower age being weakly-moderately ($R^2=0.071$ and $0.034$, respectively) associated to larger predicted values.

After applying confound mitigation approaches (feature regression or COMBAT) the partial confound test did not provide evidence of confounder bias anymore ($p > 0.05$ for all, second and third columns of Fig. \ref{fig:hcp}), neither for acquisition batch nor for age. Both feature regression and COMBAT increased the predictive performance in the case of both confounders, with COMBAT providing best performances ($R^2=0.122$ and $0.121$ when applied to remove the effect of acquisition and age, respectively).

@@

Imaging center and in-scanner motion (normalized mean framewise displacement) were statistically significantly associated with ASD diagnosis ($R^2=0.019$ and $0.028$, respectively, $p<0.001$ for both, see also Table \ref{tab:unconditional-pvals}). The model trained on the raw (unadjusted) connectivity features predicted diagnosis with a medium effect size ($R^2=0.126$, $ROC AUC = 0.71$, $p<0.001$).
The partial confounder test revealed that the raw model was significantly biased both for age group and acquisition batch (both $p<0.0001$, see first column on Fig. \ref{fig:abide}). Several sites (e.g. Carnegine Mellon University, University of Leuven, Social Brain Lab UMC Groningen) and lower motion being weakly-moderately ($R^2=0.169$ and $0.111$, respectively) associated to a lower probability for the participant to be assigned to the ASD group by the model.

Both feature regression and COMBAT seemed to significantly attenuate the effect of center, however the partial confound test still provided evidence for a significant center-bias ($p=0.04$ and $0.01$ for feature regression and COMBAT, respectively, see the second and third columns of the first row on Fig \ref{fig:abide}).

When trying to mitigate the effect of in-scanner motion (bottom row on Fig \ref{fig:abide}), feature regression obviously failed to remove the motion-bias from the model and, in fact, it introduced a paradoxical inverse dependence of the predictions on motion estimates. The partial confounder successfully detected the resulting strong bias ($p<0.0001$). Applying COMBAT to remove the effect of motion (by using binned motion estimates), in turn, seemed to effectively mitigate motion-bias, as suggested by the partial confound test ($p=0.64$, bottom right panel of Fig \ref{fig:abide}).

Both feature regression and COMBAT considerably improved the predictive performance when mitigating center-effects ($R^2=0.111$ and $0.132$ and $AUC=0.76$ and $0.75$, respectively). With both feature regression and COMBAT, however, the effort of mitigating motion effects happened at the cost of a drop in predictive performance ($R^2=0.02$ and $0.111$ and $AUC=0.59$ and $0.70$, respectively)

@@

% tab:unconditional-pvals
\renewcommand{\arraystretch}{1.2}
\begin{table}
\centering
\begin{tabular}{lll|ll|ll|ll|ll} 
dataset & conf. & method & $R^2_{y, c}$ & $p_{y, c}$ & $R^2_{\hat{y}, c}$ & $p_{\hat{y}, c}$ & $R^2_{\hat{y}, y}$ & $p_{\hat{y}, y}$ & partial test & full test  \\
\hline
HCP & acq.  & raw      & 0.032 & <0.001 & 0.071 & <0.001  & 0.095 & <0.001 & <0.0001 & <0.0001 \\
    &              & f.reg.    & & & 0.011 & 0.477 & 0.105  & <0.001  & 0.72 & <0.0001 \\
    &              & COMBAT    & & &0.013 & 0.4 & 0.122  & <0.001  & 0.88 & <0.0001\\
\hline
    & age   & raw       & 0.011 & 0.001  & 0.034 & <0.001  & 0.095  & <0.001 & <0.0001 & <0.0001 \\
    &       & f.reg.    & & &0.004 & 0.052 & 0.114  & <0.001  & 0.23 & <0.0001 \\
    &       & COMBAT    & & &0.005 & 0.048 & 0.121 & <0.001 & 0.17 & <0.0001 \\
\hline
ABIDE   & center   & raw       & 0.019  & <0.001 &  0.169 & <0.001& 0.126     & <0.001 & <0.0001 & <0.0001 \\
        &          & f.reg.    &  & &  0.042 & 0.007 & 0.173     & <0.001 & 0.04 & <0.0001 \\
        &          & COMBAT    &  & &  0.05 & 0.001 & 0.17     & <0.001 & 0.01 & <0.0001 \\
\hline
        & motion   & raw       & 0.028 & <0.001 & 0.111    &  <0.001 & 0.126    & <0.001 & <0.0001& <0.0001 \\
        &          & f.reg.    & & & 0.359    & <0.001  & 0.02   & <0.001 & <0.0001 & 0.09 \\
        &          & COMBAT    & & &  0.002 & 0.19 & 0.111     & <0.001 & 0.64 & <0.0001 \\

@@

Confounder-bias can significantly harm the generalizability and biomedical validity of predictive models. Testing confounder-bias of predictive models can be achieved by considering the "confounder-problem" as assessing \emph{conditional independence} between the predicted values and either the target or the confounder variable. These two scenarios result in two different test, testing whether the model is partially or fully driven by the given confounder.
Importantly, in predictive modelling, the model predictions often follow a non-normal distribution, thus any test for confounder bias must be, at least partly, non-parametric.
However, as shown by \citet{shah2020hardness} in their 'no free lunch' theorem, it is effectively impossible to establish a fully non-parametric conditional independence test with a valid type I error control and non-trivial power.
Here I propose the use of the conditional permutation test (CPT) of \cite{berrett2020conditional} to test confounder bias in predictive models. Adapting this approach for the confounder testing problem makes it possible to place assumptions about the joint distribution of only two out of the three variables involved; that is, it allows constructing both the partial and full confounder tests so that no assumption is made about the distribution of the model output. 
The tests are made available in the python package \emph{mlconfound} which builds upon a parallelized, computationally effective implementation of the Markov-chain Monte-Carlo based parallel pairwise sampler of \cite{berrett2020conditional}. It utilizes coefficient of determination ($R^2$ or pseudo $R^2$ in case of categorical confounder or classification) as a test statistic and assumes that the conditional distribution between the target variable and the confounder is normal. The partial and full confounder tests available in \emph{mlconfound} can be uniformly applied for any predictive model, given that realistic (e.g. cross-validated) model predictions can be obtained. Both tests require only the target variable, the model predictions and the putative confounder as input, thus, they can be performed with a negligible extra computational cost; re-fitting the model is not required.

As expected from theoretical justification, both tests displayed a valid type I error control in the simulations with the necessary assumptions, that is, even if $\yhat$ is not normally distributed. In fact, the tests were found to become somewhat conservative (i.e. lower than expected type I error rates ) in the case of high-performance predictive models (when the $R^2$ between the predicted and observed values is greater than 0.7).

@@

Next to statistical power, the practical applicability of the proposed tests is also highly affected by their robustness to violations of the normality assumptions on the conditional distribution of $\y$ and $\c$. Importantly, the simulations strongly suggest that both tests are strikingly robust to the violation of this assumption. Type I error is well controlled even in case of highly skewed and heavy tailed distributions (kurtosis > 5, skewness > 2) and gets significantly inflated only with extreme violations of normality (see the purple distribution on Fig. \ref{fig:sim-non-normal}). Due to the conservatives of the tests, if the predictive model is very accurate, even this extremely skewed and heavy tailed distribution resulted in lower than the nominal type I error rates.

@@

Validity, power and robustness renders the proposed tests as unprecedented tools for model diagnostics in a wide spectrum of predictive modelling scenarios.
For instance, in biomedical research, confounder-bias has previously often been investigated (if investigated at all) by testing the (unconditioned) association between the confounder variable and the predictive values, see e.g. \citep{spisak2020pain}. However, the significance of the unconditioned association does not necessarily imply a significant confound-bias of the model, especially if the confounder is also associated to the true target variable. In this case, namely, the model still might not be directly driven by the confounder and the dependence of the predictions on the confounder might be only marginal, to be explained solely by the confounder-target association. 
Two examples for this situation can be found in the presented analysis of real data; when applying COMBAT to mitigate age-bias in the HCP data and when regressing out the center-effect from the ABIDE data (see Table \ref{tab:unconditional-pvals}).

Arguably, in certain applications, the researcher might aim to fully eliminate confounder-related variance of the predictions and, in this case, testing unconditional independence between the confounder and the predictions might be sufficient. However, a much more realistic aim is to construct models that allow a reasonable degree of marginal confounder-association of the predictions, but are likely not directly driven by confounder-information, i.e. the prediction-confounder association is proportional to, and can likely be fully explained by, the target-confounder association.
Put simply, this is what is tested by the proposed partial confounder test.

Obviously, a true confounder bias might still be present in the model, even if the the target-confounder association likely explains the prediction-confounder association, i.e. the model might still use the confounder's effect on the features (e.g artifacts) to construct the corresponding part of the variance in the prediction. Nevertheless, in most of the cases, it does not seem likely that the model, at the same time, (i) misses that specific "part" of the target variable's variance in the features that is shared with the confounder (ii) captures exclusively that component of the confounder's variance in the features that is needed to mimic an unbiased model.
Usually a much more plausible explanation is that a direct confounder effect in the model output is simply negligible (e.g because it is not to be found in the features after confound mitigation).

@@

To demonstrate the usefulness of the proposed tests in detecting various types of confounder-bias, here they have been deployed in two research settings (a regression and a classification problem) where specific types of confounding effects have previously been described. 
First, functional connectivity data from the Human Connectome Project (HCP) \citep{van2013wu} was used to build predictive models of fluid intelligence and to test for the previously discussed confounding effect of age \citep{lohmann2021predicting, dubois2018distributed} and,  additionally, the - somewhat underdiscussed - batch-like effect of acquisition date of the data within the course of the data acquisition process.
Second, functional connectivity data from the ABIDE \citep{di2014autism} database was used to investigate the potential motion- and center-bias (as previously reported by e.g. \cite{spisak2014voxel, spisak2019optimal, gotts2013perils}) when training models that aim to predict ASD diagnosis.

@@

In case of the HCP dataset, the weak but statistically significant association between age and IQ predictions could likely exaggerate to a serious bias when testing the model on data of participants outside of the (relatively narrow)
age range of the HCP sample. In this case, the bias would likely significantly harm the out-of-sample generalizability of this model. The bias of the same model to 'acquisition batch' (phase of the study the acquisition belongs to) can also be problematic and it has not yet been thoroughly discussed in case of the HCP dataset. There can be manifold reasons for the observed association. Fluid intelligence of the participants might be, for instance, affected by a changing selection bias during participant recruitment (e.g. as a consequence of the human connectome project receiving an increasing degree of public interest during its course, the sociodemographic and educational characteristics of the volunteer sample might have changed over time). To explain the associated change in the target variable, the model might utilise changing characteristics of the imaging data (caused by e.g. slight changes in the calibration settings of the measuring equipment or the changing level of experience of the research staff). This is obviously undesirable for any brain-based predictive model of fluid intelligence.

In the ABIDE dataset, neither the medium-strong center-bias nor the age-bias is surprising but both are obviously problematic for a diagnostic model of ASD. For instance the model trained on the raw (unadjusted) features - depending on the calibration of the predicted class probabilities - might classify all participants form e.g. the CMU (Carnegie Mellon University) center as neurotypical control participants. It's easy to see that this bias can possibly lead to a very low predictive performance on out-of-center data. Similarly, a model biased by motion - next to having questionable neuroscientific validity - might systematically fail in populations with higher motion or altered motor characteristics (as known e.g. for ADHD \citep{eloyan2012automated}).

Confounder mitigation techniques, with one exception, tended to decrease confounder bias. The partial confounder test provided detailed insights into the effectiveness of the techniques for each confounder. In the HCP data, it revealed that both the acquisition-bias and the age-bias was very effectively removed by both feature regression and COMBAT (p>0.05 for all). Given the high power of the test at the sample sizes of the HCP dataset ($N=999$), any remaining confounder-bias is most probably very safely negligible and well out of the range of practical relevance.

On the other hand, the partial confoudner test also showed, that the performance of the investigated confound mitigation techniques is much less convincing for the ABIDE dataset. The center-bias of the classification in this massively multi-center dataset was, namely, found to be significant, even after feature regression and COMBAT. Nevertheless, the p-values returned by the partial confounder test were larger with two orders of magnitude than without any confound mitigation, suggesting that a large part of the center effects was still successfully removed by both approaches. While determining the relevance of the remaining bias is to be performed based on the experimental context and the requirements towards the predictive marker, significant p-values of the partial confounder test generally suggest that these models would either require more effective data harmonization or a validation on external data, that confirms the low practical relevance of the amount of bias.

Motion-bias in the ABIDE dataset was not detectable anymore after COMBAT, but remained highly significant after feature regression.
Interestingly, in the latter case, the association between the confounder and the predictions become even stronger than before but with an opposite sign. Meanwhile, the prediction performance substantially dropped (although still remained statistically significant). 
A possible explanation to these phenomena is that in-scanner motion, as previously described \citep{fournier2010motor, anzulewicz2016toward}, has manifold links to ASD and, therefore, regressing out motion estimates from the connectivity features might eliminate a significant amount of signal-of-interest (especially in functional connections that are themselves directly associated with motor deficits in ASD).
On the other hand, motion artefacts were previously shown to display complex spatiotemporal patterns in fMRI-based functional connectivity estimates \citep{spisak2014voxel} and simply regressing them out of the features is far from being sufficient for removing their effect on the model. The so-called "residual motion artefacts" and their regional interactions, as described in the referenced work, might be responsible for the paradox behavior of this model, which raises caution for using feature regression to mitigate the effect of motion artifacts and similarly complex confounders.

@@

 In sum, the application of the partial confounder test suggests that confounder-bias should be much more carefully investigated and reported in studies utilizing predictive modelling and machine learning as (i) variables as trivial as the date of the acquisition can cause significant model-bias and (ii) in certain situations, a sufficient mitigation of confounder-bias requires more effective solutions than feature regression and COMBAT. The partial confounder test can be considered as a useful objective benchmark to guide the search for a suitable confounder mitigation approach.

While the real data examples in this paper were not typical cases of the potential applications of the full confounder test, it has still provided important insights into model biases. Namely, in the case of the aforementioned paradox motion-regression model in the ABIDE dataset, the full confounder test did not provide any evidence that the model captures any extra variance over that explained by the confounder. In other words, the test accepted the null hypothesis of full model bias, suggesting that this model might be, in fact, almost exclusively driven by motion artifacts.
For all other models, the full confounder test confirmed that - even if there was a significant confounder-bias - the model was not fully driven by the confounder (in other words, the target and the prediction were not independent, given the confounders).
This examples highlights that the full confounder test might become useful in exploratory phases of model development, where models might be still severely biased and the questions is whether there is any biomedical relevant signal captured by the model, at all.

@@

The lack of rigorous statistical tests for confounder-bias significantly hampers the development of predictive models in many fields of research. Even tough several approaches for mitigating the effects of counfounder variables in machine learning have been recently proposed, these can potentially remove signal-of-interest together with the confounder signal and an informed decision about confound mitigation strategy is impossible without a proper test of confounder-bias.

To fill this critical gap in predictive model development, here I proposed two novel tests, the partial and the full confounder tests, which probe the null hypotheses of 'no confounder-bias' and 'full confounder-bias', respectively. Both approaches are based on a recently published framework for conditional permutation testing, with a mathematical justification of valid type I error control. The tests were constructed so that they do not place any assumption on the distribution of the model predictions and require only the target variable, the model predictions and the putative confounder as input, that is, they can be performed with a minimal computational overhead; without needing to re-fit the model.

Given their simplicity, high statistical power, robustness to the violation of the assumptions and thanks to the computationally effective implementation (available in python package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}}), the tests are deployable in a wide variety of applications.

As demonstrated on functional brain connectivity-based predictive models of fluid intelligence and autism, the tests can guide the optimization of the confound mitigation strategy and allows quantitative, statistical assessment of the robustness, generalizability and neurobiological validity of predictive models in biomedical research, thereby paving the way for the development of clinically useful machine learning biomarkers.
